---
title: "Bioinformatics tools to integrate and understand molecular changes associated with Immune Response, Stemness and Oncogenic processes"
author: "Antonio Colaprico, Lily Wang, & Xi Chen"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    number_sections: yes
    toc: true
vignette: >
  %\VignetteIndexEntry{Bioinformatics tools TCGAbiolinks Moonlight and DeepBlueR}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = FALSE,
  comment = "#>"
)
```

```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(knitr)
#library(papeR)
```


# Instructors

* Dr. Antonio colaprico^[Department of Public Health Sciences, Miller School of Medicine, The University of Miami, Miami, FL] (<axc1833@med.miami.edu>)
* Prof. Lily Wang^[Department of Public Health Sciences, Miller School of Medicine; McDonald Foundation Department of Human Genetics; The University of Miami, Miami, FL] (<lily.wang@med.miami.edu>)
* Prof. Steven Chen^[Department of Public Health Sciences, Miller School of Medicine; Sylvester Comprehensive Cancer Center; The University of Miami, Miami, FL] (<steven.chen@med.miami.edu>)


# Workshop Description

Recently, The Cancer Genome Atlas (TCGA’s) Pan-Cancer Atlas initiative presented a comprehensive collection of 27 studies
covering 11,000 patient tumors from 33 cancer types. These studies investigated cancer complexity from different angles by integrating multi -omics and clinical data. In particular, computational analyses have led to the identification of 299 cancer-driver genes and over 3,400 driver mutations. However, it still remains critical to clarify the consequences of each alteration and the underlying biological effects. 
We will discuss several computational tools that are useful in clarifying gene functions when performing integrative analysis of multi-omics datasets. In order to deal with the challenges of data retrieval and integration, TCGAbiolinks (Colaprico et al., 2016)
and DeepBlueR (Albrecht et al., 2017) were developed to retrieve data from TCGA, CPTAC, GTEx, GEO and IHEC, Blueprint, ENCODE, Roadmap, respectively. Tumor-specific cancer-driver-gene events and downstream impact can be elucidated with MoonlightR by integrating these datasets (Colaprico et al. 2018). TCGAbiolinks and MoolightR have been used successfully in multiple studies for oncogenic
processes identification (Ding et al., 2018), oncogenic clinically actionable driver genes discovery (Bailey et al., 2018), and comprehensive immune landscape characterization (Thorsson et al., 2018)


# Workshop: overview 

 In this workshop we will show the capability of TCGAbiolinks and Moonlight, to integrate multi -omics data from different consortium and to reproduce the six immune subtypes from TCGA PanCancer and how features (Immune Subtypes, Oncogenic Processes, Driver Genes and Stemness) can be used by the end user to expand their understating of their own un-published data. 

The workshop is organized in 4 subsections:
 
1. Data retrieval from (TCGA, GTEx, GEO and IHEC)
2. Immune Subtypes
3. Oncognic Processes and Driver Genes
4. Stemness scores
 
 
```{r, eval= FALSE, echo = FALSE,hide=TRUE, message=FALSE,warning=FALSE}
devtools::load_all(".")
```

```{r message=FALSE, warning=FALSE, include=FALSE}
library(SummarizedExperiment)
library(dplyr)
#library(DT)
```

# Data retrieval from (GDC, TCGA, GEO and IHEC)

You can easily query - download - prepare multi -omics data from GDC:
. Gene expression
. Copy number
. Protein expression (RRPA)
. Methylation 
. Clinical data
. microRNA 

```{r message=FALSE, warning=FALSE, include=FALSE}
library(SummarizedExperiment)
library(dplyr)
#library(DT)
```

## GDC TCGA Gene Expression data (IlluminaHiSeq_RNASeqV2) using TCGAbiolinks

You can easily search TCGA samples, download and prepare a matrix of gene expression.
```{r, eval = FALSE}

#DataDir <- "~/Dropbox (Personal)/Umiami/TCGAanalysis/GDCdata/TCGA-BRCA"

# Firstly we install TCGAbiolinks from bioconductor or from github

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("TCGAbiolinks")

require(TCGAbiolinks)


cancerType <- "BRCA"

# Query platform Illumina HiSeq with a list of barcode 
query <- GDCquery(project = paste0("TCGA-",cancerType),
                  data.category = "Gene expression",
                  data.type = "Gene expression quantification",
                  experimental.strategy = "RNA-Seq",
                  platform = "Illumina HiSeq",
                  file.type = "results",
                  legacy = TRUE)

# We select 10 tumor and 10 normal samples
Sample_sel <- query$results[[1]]$cases

Sample_sel_TP <- TCGAquery_SampleTypes(barcode = Sample_sel,typesample = "TP")
Sample_sel_NT <- TCGAquery_SampleTypes(barcode = Sample_sel,typesample = "NT")                               

Sample_sel_short <- c(Sample_sel_TP[1:10],Sample_sel_NT[1:10])

# we need to create a new query with the selected barcodes
query_down <- GDCquery(project = paste0("TCGA-",cancerType),
                  data.category = "Gene expression",
                  data.type = "Gene expression quantification",
                  experimental.strategy = "RNA-Seq",
                  platform = "Illumina HiSeq",
                  file.type = "results",
                  barcode = Sample_sel_short,
                  legacy = TRUE)


# Download a list of barcodes with platform IlluminaHiSeq_RNASeqV2
GDCdownload(query_down,directory = DataDir)

# Prepare expression matrix with geneID in the rows and samples (barcode) in the columns
# rsem.genes.results as values
BRCARnaseqSE <- GDCprepare(query_down,directory = DataDir)
# 
# For gene expression if you need to see a boxplot correlation and AAIC plot to define outliers you can run
dataPrep <- TCGAanalyze_Preprocessing(BRCARnaseqSE)
```

The result is shown below:

```{r, eval = TRUE, echo = FALSE,size = 8}
library(TCGAbiolinks)
dataGE <- dataBRCA[sample(rownames(dataBRCA),10),sample(colnames(dataBRCA),7)]

knitr::kable(dataGE[1:10,2:3], digits = 2, 
             caption = "Example of a matrix of gene expression (10 genes in rows and 2 samples in columns)",
             row.names = TRUE)
```

The result from TCGAanalyze_Preprocessing is shown below:
```{r, fig.width=6, fig.height=4, echo=FALSE, fig.align="center"}
library(png)
library(grid)
img <- readPNG("PreprocessingOutput.png")
grid.raster(img)
```

## GDC TCGA Protein Expression data (RPPA) using TCGAbiolinks

You can easily search TCGA samples, download and prepare a matrix of protein RPPA expression.
```{r, eval = FALSE}

cancerType <- "BRCA"

query.RPPA <- GDCquery(project = paste0("TCGA-",cancerType), 
                       legacy = TRUE,
                       data.category = "Protein expression",
                       platform = "MDA_RPPA_Core", 
                       data.type = "Protein expression quantification",
                       file.type = "expression", 
                       sample.type = c("Primary solid Tumor"))

samples.RPPA <- query.RPPA$results[[1]]$cases


query.RPPA.down <- GDCquery(project = paste0("TCGA-",cancerType), 
                            legacy = TRUE,
                            data.category = "Protein expression",
                            data.type = "Protein expression quantification",
                            platform = "MDA_RPPA_Core", 
                            file.type = "expression", 
                            sample.type = c("Primary solid Tumor"),
                            barcode = samples.RPPA)


GDCdownload(query.RPPA.down, 
            directory = PathDir)

data.RPPA <- GDCprepare(query.RPPA.down, 
                        directory = PathDir)

```

## GDC TCGA Copy number variation (CNV) Affymetrix SNP Array 6.0 using TCGAbiolinks

You can easily search TCGA samples, download and prepare a matrix of CN segments.
```{r, eval = FALSE}
require(TCGAbiolinks)

cancerType <- "BRCA"
query.cnv <- GDCquery(project = cancerType,
                      data.category =  "Copy number variation",
                      legacy = TRUE,
                      data.type = "Copy number segmentation",
                      platform = "Affymetrix SNP Array 6.0",
                      file.type = "nocnv_hg19.seg",
                      sample.type = c("Primary solid Tumor"))

samples.cnv <- query.cnv$results[[1]]$cases[1:20]

query.cnv.down <- GDCquery(project = cancerType,
                      data.category =  "Copy number variation",
                      legacy = TRUE,
                      platform = "Affymetrix SNP Array 6.0",
                      data.type = "Copy number segmentation",
                      file.type = "nocnv_hg19.seg",
                      sample.type = c("Primary solid Tumor"),
                      barcode = samples.cnv)

GDCdownload(query.cnv.down,
            directory = PathDir)

data.cnv <- GDCprepare(query.cnv.down, 
                        directory = PathDir)

data.cnv <- as.data.frame(data.cnv)
```

## GDC TCGA DNA methylation Illumina Human Methylation 450 using TCGAbiolinks

You can easily search TCGA samples, download and prepare a matrix of DNA methylation probes.
```{r, eval = FALSE}
require(TCGAbiolinks)
cancerType <- "BRCA"
query.met <- GDCquery(project = cancerType, 
                      legacy = TRUE,
                      data.category = "DNA methylation",
                      platform = "Illumina Human Methylation 450",
                      sample.type = c("Primary solid Tumor"))

samples.met <- query.met$results[[1]]$cases[1:20]

query.met.down <- GDCquery(project = cancerType, 
                      legacy = TRUE,
                      data.category = "DNA methylation",
                      platform = "Illumina Human Methylation 450",
                      sample.type = c("Primary solid Tumor"),
                      barcode = samples.met)

GDCdownload(query.met.down,
            directory = PathDir)

data.met <- GDCprepare(query.met.down, 
                       directory = PathDir)

```

## GDC TCGA Mutation MAF using TCGAbiolinks

You can easily search TCGA samples, download and prepare a matrix of Mutations.
```{r, eval = FALSE}
cancerType <- "TCGA-BRCA"
query.mut <- GDCquery(project = cancerType, 
                      data.category = "Simple nucleotide variation", 
                      data.type = "Simple somatic mutation",
                      access = "open", 
                      sample.type = c("Primary solid Tumor"),
                      legacy = TRUE)

# Check maf availables
query.mut$results[[1]]$file_name

query.mut <- GDCquery(project = cancerType, 
                      data.category = "Simple nucleotide variation", 
                      data.type = "Simple somatic mutation",
                      access = "open", 
                      sample.type = c("Primary solid Tumor"),
                      legacy = TRUE,
                      file.type = query.mut$results[[1]]$file_name[1])

samples.mut <- query.mut$results[[1]]$cases


query.mut.down <- GDCquery(project = cancerType, 
                           data.category = "Simple nucleotide variation", 
                           data.type = "Simple somatic mutation",
                           access = "open", 
                           sample.type = c("Primary solid Tumor"),
                           legacy = TRUE,
                           barcode = samples.mut,
                           file.type = query.mut$results[[1]]$file_name[1])

GDCdownload(query.mut.down,
            directory = PathDir)

data.mut <- GDCprepare(query.mut.down, 
                       directory = PathDir)


```


## GDC TCGA microRNA Expression using TCGAbiolinks

You can easily search TCGA samples, download and prepare a matrix of microRNA Expression
```{r, eval = FALSE}
cancerType <- "TCGA-BRCA"
query.miR <- GDCquery(project = cancerType, 
                      data.category = "Gene expression",
                      data.type = "miRNA gene quantification",
                      platform = "Illumina HiSeq",
                      file.type = "hg19.mirbase20.mirna.quantification",
                      legacy = TRUE)
samples.miR <- query.miR$results[[1]]$cases

query.miR.down <- GDCquery(project = cancerType, 
                      data.category = "Gene expression",
                      data.type = "miRNA gene quantification",
                      platform = "Illumina HiSeq",
                      file.type = "hg19.mirbase20.mirna.quantification",
                      legacy = TRUE,
                      barcode = TNBCsamplesmiRlong)

GDCdownload(query.miR.down,
            directory = PathDir)

data.miR <- GDCprepare(query.miR.down, 
                       directory = PathDir)

```

## GDC TCGA Clinical Data using TCGAbiolinks

You can easily search TCGA samples, download and prepare a matrix of clinical data
```{r, eval = FALSE}
cancerType <- "BRCA"
dataClin <- GDCquery_clinic(project = paste0("TCGA-",cancerType),type = "clinical")
```

## GTEx Data using TCGAbiolinks

You can easily search GTEx samples, download and prepare a matrix of gene expression.
```{r, eval = FALSE}
data_gtex_brain <- TCGAquery_recount2("gtex", tissue = "brain")

```

## IHEC gene expression data using DeepBlueR

You can easily search IHEC Blueprint samples, download and prepare a matrix of gene expression.
```{r, eval = FALSE}


require(DeepBlueR)
require(dplyr)


# List all BLUEPRINT samples
blueprint_samples <- deepblue_list_samples(
    extra_metadata = list("source" = "BLUEPRINT Epigenome"))

# Extract their ids
blueprint_samples_ids <- deepblue_extract_ids(blueprint_samples)

# Select gene expression data. We assign gene names using Gencode 22
gene_exprs_query <- deepblue_select_expressions(sample_ids = blueprint_samples_ids,
                                                expression_type = "gene",
                                                gene_model = "gencode v22")

gene_exprs_query <- deepblue_select_expressions(sample_ids = blueprint_samples_ids,
                                                expression_type = "gene",
                                                gene_model = "gencode v19")



# We request the data and define the output format
request = deepblue_get_regions(query_id = gene_exprs_query,
                               "@GENE_ID(gencode v19),FPKM,@BIOSOURCE,@SAMPLE_ID")

# We download the data
gene_regions <- deepblue_download_request_data(request)

# We retain a table mapping sample ids to bisources
sample_names <- dplyr::select(gene_regions, `@BIOSOURCE`, `@SAMPLE_ID`) %>%
    dplyr::distinct()

# We filter out duplicated gene entries
genes_one_sample <- dplyr::filter(gene_regions, `@SAMPLE_ID` == "s10678")
duplicated_genes <- genes_one_sample[
    which(duplicated(genes_one_sample$`@GENE_ID(gencode v22)`)),
    "@GENE_ID(gencode v22)"]


genes_one_sample <- dplyr::filter(gene_regions, `@SAMPLE_ID` == "s10678")
duplicated_genes <- genes_one_sample[
    which(duplicated(genes_one_sample$`@GENE_ID(gencode v19)`)),
    "@GENE_ID(gencode v19)"]


# We convert the gene expression from a list to a data frame and subsequently...
genes_matrix = dplyr::filter(gene_regions,
                             !(`@GENE_ID(gencode v22)` %in% duplicated_genes)) %>%
    dplyr::select(-`@BIOSOURCE`) %>%
    tidyr::spread(key = `@SAMPLE_ID`, value = FPKM)

# ...to a numeric matrix
genes <- genes_matrix[,1]
genes_matrix <- data.matrix(genes_matrix[,-1])
rownames(genes_matrix) <- genes

### OUTPUT
### genes_matrix : The gene expression matrix for all 276 BLUEPRINT samples
### sample_names : A mapping table from sample id to cell type / biosource

save(genes_matrix, file = "IHEC_genes_matrix.Rdata")
save(sample_names, file = "IHEC_Sample_names.rdata")

```

## GEO gene expression data using MoonlightR and GEOquery

## Preparing Gene Expression for GEO (platform and GSE) for Nazor et al.,  Cell Stem Cell 2012 

You can easily search GEO samples, download and prepare a matrix of gene expression.
```{r, eval = FALSE}
require(MoonlightR)
library(devtools)
install_github("ibsquare/MoonlightR") 

dataNazor <- getDataGEO(GEOobject = "GSE30652",
                                    platform =  "GPL6947")

require(SummarizedExperiment)
GSE30652 <- as.data.frame(exprs(dataNazor))
GSE30652_non_norm <- cbind(ILMN = rownames(GSE30652),
                           IDmean = rowMeans(GSE30652),
                           GSE30652)



dataNazor_samples <- pData(dataNazor)
dataNazor_samples <- as.data.frame(dataNazor_samples)
dataNazor_samples <- subset(dataNazor_samples,
                              select = c("geo_accession","characteristics_ch1.2"))
colnames(dataNazor_samples)[2] <- "CellType"

dataNazor_samples$CellType <- gsub("cell type: ","",dataNazor_samples$CellType)
dataNazor_samples$CellType <- gsub(", undifferentiated","",dataNazor_samples$CellType)

GPL6947_13512 <- fData(dataNazor)
GPL6947_13512_annot <- as.data.frame(GPL6947_13512)
GPL6947_13512_annot <- subset(GPL6947_13512_annot,
                              select = c("ID","Gene.symbol"))



GSE30652_merge <- merge(x = GPL6947_13512_annot,
                        y = GSE30652_non_norm,
                        by.x = "ID",
                        by.y = "ILMN")


GSE30652_merge <- GSE30652_merge[order(GSE30652_merge$IDmean,decreasing = TRUE),]
GSE30652_merge <- GSE30652_merge[!duplicated(GSE30652_merge$Gene.symbol),]
NazorMatrix <- GSE30652_merge
rownames(NazorMatrix) <- NazorMatrix$Gene.symbol

NazorMatrix <- NazorMatrix[,dataNazor_samples$geo_accession]



```
# Stemness Analysis

## Validation of Stemness signature in external dataset. Nazor et al.,  Cell Stem Cell 2012 
```{r, eval = FALSE}
# Firstly we have previously prepared the Gene Expression matrix (genes in rows , samples in columns) 
# for Nazor's dataset using Moonlight and getGEO's function.

TCGA_mRNA_StemScoreTable <- TCGAanalyze_Stemness(stemSig = PCBC_stemSig,
                                          dataGE = NazorMatrix)



tab_mRNASi <- TCGA_mRNA_StemScoreTable

tab_mRNASi_merged <- merge(x = tab_mRNASi,
                    y = dataNazor_samples,
                    by.x = "Sample",
                    by.y = "geo_accession")
require(ggplot2)
require(ggpubr)

colnames(tab_mRNASi_merged)[3] <- "mRNASi"

tab_mRNASi_merged[tab_mRNASi_merged$CellType %in% "embryonic stem cell","CellType"] <- "ES"
tab_mRNASi_merged[tab_mRNASi_merged$CellType %in% "induced pluripotent stem cell","CellType"] <- "IPS"
tab_mRNASi_merged[tab_mRNASi_merged$CellType %in% "parthenogentic embryonic stem cell","CellType"] <- "Parthenogentic"
tab_mRNASi_merged[tab_mRNASi_merged$CellType %in% "Somatic.Primary","CellType"] <- "Primary"
tab_mRNASi_merged[tab_mRNASi_merged$CellType %in% "Somatic.Tissue","CellType"] <- "Tissue"

tab_mRNASi_merged <- tab_mRNASi_merged[order(tab_mRNASi_merged$mRNASi, decreasing = FALSE),]
library(forcats)
p <- ggplot(data=tab_mRNASi_merged, aes(x=Sample, y=mRNASi, fill = CellType)) +
    geom_bar(stat="identity")+
    scale_colour_gradient2()+
    #coord_flip()+
   # ylim(0, 15)+
    #scale_x_discrete(limits = df1$Tissue)+
    theme_classic() +
    theme(axis.title.x=element_blank(),
                          axis.text.x=element_blank(),
                          axis.ticks.x=element_blank())+
    scale_fill_manual("legend", values = c("Primary" = "orange",
                                           "Parthenogentic" = "blue",
                                           "Tissue" = "darkgreen",
                                           "ES" = "red",
                                           "IPS" = "pink"))
p <- p + theme(legend.position="bottom")

p <- p + aes(x = fct_inorder(Sample))
ggsave(p, file = "Validation_mRNASi_Nazor.png", width = 6,height = 6)


```


The result from TCGAanalyze_Stemness validation in Nazor's Dataset is shown below:
```{r, fig.width=6, fig.height=4, echo=FALSE, fig.align="center"}
library(png)
library(grid)
img <- readPNG("Validation_mRNASi_Nazor.png")
grid.raster(img)
```


## Validation of Stemness signature in external dataset. IHEC  Author et al.,  Cell 2016 
```{r, eval = FALSE}

# stemness score with IHEC samples
load("./IHEC_genes_matrix.Rdata")
load("./IHEC_Sample_names.rdata")
IHECMatrix <- genes_matrix
require(stringr)
IHEC_annot <- sample_names
colnames(IHEC_annot) <- c("Diffname_short","X1")
IHEC_annot <- as.data.frame(IHEC_annot)

library("biomaRt")
ensemblsIDS <- str_split_fixed(rownames(IHECMatrix),"\\.",2)[,1]
rownames(IHECMatrix)<- ensemblsIDS

IHEC_annot_common <- IHEC_annot[IHEC_annot$X1 %in% colnames(IHECMatrix),]


# mapping Ensembl IDs to gene Symbol

require(TCGAbiolinks)
library(clusterProfiler)
library(org.Hs.eg.db)
gene.df <- bitr(ensemblsIDS, fromType = "ENSEMBL",
                toType = c( "ENTREZID", "SYMBOL"),
                OrgDb = org.Hs.eg.db)


IHECMatrix_sel <- IHECMatrix[gene.df$ENSEMBL,]
rownames(IHECMatrix_sel) <-gene.df$SYMBOL
IHECMatrix_sel <- as.data.frame(IHECMatrix_sel)

IHECMatrix_sel_filt <- TCGAanalyze_Filtering(tabDF = IHECMatrix_sel,
                                             method = "quantile",
                                             qnt.cut = 0.25)

TCGA_mRNA_StemIHEC <- TCGAanalyze_Stemness(stemSig = PCBC_stemSig,
                                                 dataGE = IHECMatrix_sel_filt)

colnames(TCGA_mRNA_StemIHEC)[3] <- "mRNASi"
colnames(IHEC_annot)[1] <- "CellType"
colnames(IHEC_annot)[2] <- "Sample"
TCGA_mRNA_StemIHEC_merge<- merge(x = TCGA_mRNA_StemIHEC,
                                 y = IHEC_annot,
                                 by.x = "Sample",
                                 by.y = "Sample")
tab_mRNASi_merged <- TCGA_mRNA_StemIHEC_merge

CellType <- table(tab_mRNASi_merged$CellType)



CellType_filt <- CellType[CellType > 5]

tab_mRNASi_merged_filt <- tab_mRNASi_merged[tab_mRNASi_merged$CellType %in% names(CellType_filt), ]
tab_mRNASi_merged <- tab_mRNASi_merged_filt


tab_mRNASi_merged <- tab_mRNASi_merged[order(tab_mRNASi_merged$mRNASi, decreasing = FALSE),]
require(ggplot2)
library(forcats)
p <- ggplot(data=tab_mRNASi_merged, aes(x=Sample, y=mRNASi, fill = CellType)) +
    geom_bar(stat="identity")+
    scale_colour_gradient2()+
    #coord_flip()+
    # ylim(0, 15)+
    #scale_x_discrete(limits = df1$Tissue)+
    theme_classic() +
    theme(axis.title.x=element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank())
p <- p + theme(legend.position="right")

p <- p + aes(x = fct_inorder(Sample))
p <- p + theme(text = element_text(size=14))
ggsave(p, file = "Validation_mRNASi_IHEC.png", width = 12,height = 6)
```


The result from TCGAanalyze_Stemness validation in IHEC's Dataset is shown below:
```{r, fig.width=6, fig.height=4, echo=FALSE, fig.align="center"}
library(png)
library(grid)
img <- readPNG("Validation_mRNASi_IHEC.png")
grid.raster(img)
```


## Generating Stemness score for TCGA gene expression data (mRNASi). Malta et al. Cell, 2018 

```{r, eval = FALSE}
# Firstly we load Gene Expression matrix (genes in rows , samples in columns) 
# from a previous generated pancancer gene expression matrix
load("~/Dropbox (Personal)/Umiami/TCGAanalysis/Stemness/dataFilt_panCancer33new.Rdata")

curCancer <- "BRCA"
# We have previously generated a table with 33 cancer types barcodes and molecular subtypes.

dataSubt_PanCancer <- PanCancerAtlas_subtypes()

# Selecting TCGA breast cancer

dataSubt_curCancer <- dataSubt_PanCancer[dataSubt_PanCancer$cancer.type %in% "BRCA",]

commonSamples <- intersect(dataSubt_curCancer$pan.samplesID, colnames(dataFilt))

dataFilt_curCancer <- dataFilt[,commonSamples]
load("~/Dropbox (Personal)/Umiami/Github/TCGAbiolinks/data/PCBC_stemSig.rda")

TCGA_mRNA_StemScoreTable <- TCGAanalyze_Stemness(stemSig = PCBC_stemSig,
                                          dataGE = dataFilt_curCancer)

colnames(TCGA_mRNA_StemScoreTable)[1] <- "barcode"

sampleNT <- TCGAquery_SampleTypes(barcode = TCGA_mRNA_StemScoreTable$barcode,typesample = "NT")
sampleTP <- TCGAquery_SampleTypes(barcode = TCGA_mRNA_StemScoreTable$barcode,typesample = "TP")
#sampleTM <- TCGAquery_SampleTypes(barcode = TCGA_mRNA_StemScoreTable$barcode,typesample = "TM")
#sampleTAM <- TCGAquery_SampleTypes(barcode = TCGA_mRNA_StemScoreTable$barcode,typesample = "TAM")
#sampleTB <- TCGAquery_SampleTypes(barcode = TCGA_mRNA_StemScoreTable$barcode,typesample = "TB")

rownames(TCGA_mRNA_StemScoreTable) <- TCGA_mRNA_StemScoreTable$barcode

TCGA_mRNA_StemScoreTable[sampleNT,"SampleType"] <- "NT"
TCGA_mRNA_StemScoreTable[sampleTP,"SampleType"] <- "TP"
#TCGA_mRNA_StemScoreTable[sampleTM,"SampleType"] <- "TM"
#TCGA_mRNA_StemScoreTable[sampleTAM,"SampleType"] <- "TM"
#TCGA_mRNA_StemScoreTable[sampleTB,"SampleType"] <- "TB"

tab_mRNASi <- TCGA_mRNA_StemScoreTable

tab_mRNASi_merged <- merge(x = tab_mRNASi,
                    y = dataSubt_PanCancer,
                    by.x = "barcode",
                    by.y = "pan.samplesID")
require(ggplot2)
require(ggpubr)

colnames(tab_mRNASi_merged)[3] <- "mRNASi"



p<-ggplot(tab_mRNASi_merged, aes(x=SampleType, y=mRNASi, fill=SampleType))
p <- p + theme_classic()

p <- p +  theme(legend.position="none")
p <- p + rotate_x_text(45)
p <- p + geom_jitter(shape=16, position=position_jitter(0.2), color = "black")
p <- p + geom_boxplot(position=position_dodge(1),outlier.colour = NA)
p <- p + theme(text = element_text(size=20))
ggsave(p , filename = "TCGA_BRCA_mRNASi_TP_NT.png",width = 5,height = 6)


tab_mRNASi_merged_TP <- tab_mRNASi_merged[tab_mRNASi_merged$SampleType %in% "TP",]

p<-ggplot(tab_mRNASi_merged_TP, aes(x=Subtype_Selected, y=mRNASi, fill=Subtype_Selected))
p <- p + theme_classic()

p <- p +  theme(legend.position="none")
p <- p + rotate_x_text(45)
p <- p + geom_jitter(shape=16, position=position_jitter(0.2), color = "black")
p <- p + geom_boxplot(position=position_dodge(1),outlier.colour = NA)
p <- p + theme(text = element_text(size=20))
ggsave(p , filename = "TCGA_BRCA_mRNASi_subtypes.png",width = 5,height = 6)



```

The result from TCGAanalyze_Stemness is shown below:
```{r, fig.width=6, fig.height=4, echo=FALSE, fig.align="center"}
library(png)
library(grid)
img <- readPNG("TCGA_BRCA_mRNASi_TP_NT.png")
grid.raster(img)
```

The result from TCGAanalyze_Stemness with molecular subtypes is shown below:
```{r, fig.width=6, fig.height=4, echo=FALSE, fig.align="center"}
library(png)
library(grid)
img <- readPNG("TCGA_BRCA_mRNASi_subtypes.png")
grid.raster(img)
```


## Generating Stemness score for TCGA LGG and GBM gene expression data (mRNASi). Malta et al. Cell, 2018 

```{r, eval = FALSE}

# working with LGG and GBM
curCancer <- c("LGG","GBM")
# We have previously generated a table with 33 cancer types barcodes and molecular subtypes.
require(TCGAbiolinks)
dataSubt_PanCancer <- PanCancerAtlas_subtypes()

# Selecting TCGA breast cancer

dataSubt_curCancer <- dataSubt_PanCancer[dataSubt_PanCancer$cancer.type %in% curCancer,]

sampleCurCancer <- colnames(dataFilt)
sampleCurCancer <- sampleCurCancer[substr(sampleCurCancer,1,12) %in% dataSubt_curCancer$pan.samplesID]

dataFilt_curCancer <- dataFilt[,sampleCurCancer]
load("~/Dropbox (Personal)/Umiami/Github/TCGAbiolinks/data/PCBC_stemSig.rda")

TCGA_mRNA_StemScoreTable <- TCGAanalyze_Stemness(stemSig = PCBC_stemSig,
                                                 dataGE = dataFilt_curCancer)

colnames(TCGA_mRNA_StemScoreTable)[1] <- "barcode"


tab_mRNASi <- TCGA_mRNA_StemScoreTable
tab_mRNASi <- cbind(barcode12 = substr(tab_mRNASi$barcode,1,12),
                    tab_mRNASi)
tab_mRNASi$barcode12 <- as.character(tab_mRNASi$barcode12)


tab_mRNASi_merged <- merge(x = tab_mRNASi,
                           y = dataSubt_PanCancer,
                           by.x = "barcode12",
                           by.y = "pan.samplesID")
require(ggplot2)
require(ggpubr)

colnames(tab_mRNASi_merged)[4] <- "mRNASi"

p<-ggplot(tab_mRNASi_merged, aes(x=Subtype_Selected, y=mRNASi, fill=Subtype_Selected))
p <- p + theme_classic()
p <- p +  theme(legend.position="none")
p <- p + rotate_x_text(45)
p <- p + geom_jitter(shape=16, position=position_jitter(0.2), color = "black")
p <- p + geom_boxplot(position=position_dodge(1),outlier.colour = NA)
p <- p + theme(text = element_text(size=20))
ggsave(p , filename = "TCGA_LGG_GBM_mRNASi_subtypes.png",width = 5,height = 6)
```


The result from TCGAanalyze_Stemness for LGG and GBM is shown below:
```{r, fig.width=6, fig.height=4, echo=FALSE, fig.align="center"}
library(png)
library(grid)
img <- readPNG("TCGA_LGG_GBM_mRNASi_subtypes.png")
grid.raster(img)
```


## Stemness index is associated with overall survival (Glioma example)

```{r, eval = FALSE}

#working with survival glioma

tab_mRNASi_merged <- cbind(mRNASi_level = rep("mRNASi mod",nrow(tab_mRNASi_merged)),
                           tab_mRNASi_merged)

tabSi<- tab_mRNASi_merged
tabSi$mRNASi_level <- as.character(tabSi$mRNASi_level)

tabSi[tabSi$mRNASi < quantile(tabSi$mRNASi,1/3),"mRNASi_level"] <- "mRNASi Low"
tabSi[tabSi$mRNASi > quantile(tabSi$mRNASi,2/3),"mRNASi_level"] <- "mRNASi High"


require(TCGAbiolinks)

dataClin_LGG <- GDCquery_clinic(project = "TCGA-LGG",type = "clinical")
dataClin_GBM <- GDCquery_clinic(project = "TCGA-GBM",type = "clinical")

dataClin_LGG_GBM <- rbind(dataClin_LGG,dataClin_GBM)

dataClin_LGG_GBM <- dataClin_LGG_GBM[dataClin_LGG_GBM$submitter_id %in% tabSi$barcode12,]

dataClin_merged <- merge(x = dataClin_LGG_GBM,
                         y = tabSi,
                         by.x = "submitter_id",
                         by.y = "barcode12")




p <- TCGAanalyze_survival(dataClin_merged,
                          clusterCol = "mRNASi_level",
                          conf.int = FALSE,
                          main = "TCGA LGG GBM mRNASi", 
                          height = 10,
                          width=10,
                          risk.table = TRUE,
                          filename = NULL)

p <- p$plot

p <- p  + theme(legend.position = "bottom")

ggsave(p , filename = "TCGA_LGG_GBM_mRNASi_survival.png",width = 5,height = 6)
```


The result from the Stemness-survival association for LGG and GBM is shown below:
```{r, fig.width=6, fig.height=4, echo=FALSE, fig.align="center"}
library(png)
library(grid)
img <- readPNG("TCGA_LGG_GBM_mRNASi_survival.png")
grid.raster(img)
```

# Immune subtypes

In this section we generate the Immune subtypes for TCGA and GEO tumors.
Figure1C readapted from Thorsson et al., Immunity, 2018, to summarize features of six different immune subtypes.

```{r, fig.width=6, fig.height=4, echo=FALSE, fig.align="center"}
library(png)
library(grid)
img <- readPNG("ImmuneLandscape_Fig1B.png")
grid.raster(img)
```


## KM survival analysis of TCGA KIRC samples with Immune Subtypes from Thorsson et al., Immunity, 2018
```{r, eval = FALSE}
download.file(url = "https://ars.els-cdn.com/content/image/1-s2.0-S1074761318301213-mmc2.xlsx",
               destfile = "X1_s2_0_S1074761318301213_mmc2")
X1_s2_0_S1074761318301213_mmc2 <- read_excel("X1_s2_0_S1074761318301213_mmc2")
X1_s2_0_S1074761318301213_mmc2 <- as.data.frame(X1_s2_0_S1074761318301213_mmc2)

CancerType <- c("KIRC")

ImmuneSubtypes <- as.data.frame(X1_s2_0_S1074761318301213_mmc2)

ImmuneSubtypes <- ImmuneSubtypes[ImmuneSubtypes$`TCGA Study` %in% CancerType,]
ImmuneSubtypes <- ImmuneSubtypes[ImmuneSubtypes$`Immune Subtype`!="NA",]

dataClin_KIRC <- GDCquery_clinic(project = "TCGA-KIRC",type = "clinical")

dataClin_merged <- merge(x = dataClin_KIRC,
                         y = ImmuneSubtypes,
                         by.x = "submitter_id",
                         by.y = "TCGA Participant Barcode")


p <- TCGAanalyze_survival(dataClin_merged,
                          clusterCol = "Immune Subtype",
                          conf.int = FALSE,
                          main = "TCGA KIRC Immune Subtypes",
                          height = 10,
                          width=10,
                          risk.table = TRUE,
                          filename = NULL)

p <- p$plot

p <- p  + theme(legend.position = "right")

ggsave(p , filename = "TCGA_KIRC_ImmuneSubtypes_survival.png",width = 10,height = 6)

```

The result from the Immune subtypes -survival association for BRCA samples is shown below:
```{r, fig.width=6, fig.height=4, echo=FALSE, fig.align="center"}
library(png)
library(grid)
img <- readPNG("TCGA_KIRC_ImmuneSubtypes_survival.png")
grid.raster(img)
```

## Immune Subtypes for GEO samples (KIRC example) using data from Jones et al.,Clin Cancer Res 2005  (16115910)

```{r, eval = FALSE}

require(MoonlightR)
 
#Jones J, Otu H, Spentzos D, Kolia S et al. Gene signatures of progression and metastasis in renal cell cancer. 
# Clin Cancer Res 2005 Aug 15;11(16):5730-9. PMID: 16115910

dataGEO_KIRC<- getDataGEO(GEOobject = "GSE15641",
                        platform =  "GPL96")
require(SummarizedExperiment)

GSE15641 <- as.data.frame(exprs(dataGEO_KIRC))
GSE15641_non_norm <- cbind(ILMN = rownames(GSE15641),
                           IDmean = rowMeans(GSE15641),
                           GSE15641)

GSE15641_annot <- fData(dataGEO_KIRC)
GSE15641_annot <- as.data.frame(GSE15641_annot)
GSE15641_annot <- subset(GSE15641_annot,
                              select = c("ID","Gene.symbol"))


dataGEO_samples <- pData(dataGEO_KIRC)
dataGEO_samples <- as.data.frame(dataGEO_samples)
dataGEO_samples <- subset(dataGEO_samples,
                            select = c("geo_accession","source_name_ch1"))
colnames(dataGEO_samples)[2] <- "CellType"

dataGEO_samples <- dataGEO_samples[dataGEO_samples$CellType %in% "cancerous human kidney tissue, clear cell RCC",]

GSE15641_merge <- merge(x = GSE15641_annot,
                        y = GSE15641_non_norm,
                        by.x = "ID",
                        by.y = "ILMN")


GSE15641_merge <- GSE15641_merge[order(GSE15641_merge$IDmean,decreasing = TRUE),]
GSE15641_merge <- GSE15641_merge[!duplicated(GSE15641_merge$Gene.symbol),]
GSE15641_Matrix <- GSE15641_merge
rownames(GSE15641_Matrix) <- GSE15641_Matrix$Gene.symbol

GSE15641_Matrix <- GSE15641_Matrix[,dataGEO_samples$geo_accession]

dataImmuneSubtype <- TCGAanalyze_ImmuneSubtypes(ImmuneMW = ImmuneMW,
                                                dataGE = GSE15641_Matrix)

```


# Oncogenic processes and driver genes using MoonlightR

To include some examples from
http://bioconductor.org/packages/release/bioc/vignettes/MoonlightR/inst/doc/Moonlight.html



# Session Information
******
```{r sessionInfo}
sessionInfo()
```

